Boilerplate logger

### 2020-05-03 paper/paper.tex
[paper] Updates from review for homework 6

\title{Boilerplate logger: A logger for developers by developers}
\subtitle{Homework 6}
\author{Evan Lee}
\affiliation{%
    \institution{University of Colorado Boulder}
    \city{Boulder}
    \state{Colorado}
}
\email{evan.n.lee@colorado.edu}
\date{April 2020}

@@

@@

\begin{abstract}
    The logging aspect of any software project always incurs a non-zero developer investment, both from a development and testing perspective as well as a operational and error tracing perspective. A significant portion of the thought and energy behind logging decisions arise from nearly-universal problems and use cases and yet still require each distinct developer to undergo the same rote process of decision making. Even with these factors and costs in play, logging is typically not considered as a first-class issue which leads to hidden cost in terms of development cycles or unnecessary iteration to assist in tracing and debugging. After having identified a widely realized use case of logging in dev-test cycles, we introduced and defined the concept of "boilerplate logging" and developed a python logging library that largely abstracts the human component to logging decisions, simultaneously solving the hidden up-front development cost in logging decision making as well as eliminating any future costs in iterating on reactive postmortem logging.
\end{abstract}

@@

    % What is our contribution?
    Our contribution to this effort is two-fold: a collection and analysis of log usage in widely-used open-source libraries as well as an implementation that displays the ideas and positions presented below. Popular open-source libraries and tools are of particular interest in this space as not only do they have a wide range of contributors, but there is a communal sense of "being a good neighbor" when writing open-source code; particular care is given to the logging statements that are eventually committed and merged into the mainstream base and thus particular thought and effort is given to the balancing act between verbosity and utility.

@@

    In order to assist the reader in visualizing the analysis and ideas that this paper presents as well as the benefits of implementation, we will cover a code example from a widely-used python library for interacting with SOAP interfaces: \lstinline{python-zeep}. The code snippet can be found in Listing \ref{lst:explosion-example}\footnote{Source: \href{https://github.com/mvantellingen/python-zeep/blob/4df383021e31372c111bc26cbf2e4535deaee04e/src/zeep/transports.py\#L47}{python-zeep/transports.py}, lines 47-84. (sha: 4df383021e31372c111bc26cbf2e4535deaee04e)}. Reading through this 31-line snippet, most of the lines written by the developer deal with logging in one way or another, whether that is making logging decisions (lines 2-5, 12-22) or dealing with emitting log messages themselves (lines 6, 24). In fact, only three lines (lines 8-10) deal with any actual functionality that this method is meant to drive.

@@

    \begin{quote}
    HTTP Post to https://api.colorado.edu/: {"test": "test"}
    \end{quote}

@@

    \begin{quote}
        [transports.post]: [args: ('https://api.colorado.edu/', \{"test": "test"\}, \{\})] [kwargs: \{\}]
    \end{quote}

@@

    The key distinction between boilerplate logging and that of other standard logging libraries is a matter of control. In general, standard third-party and language-specific logging libraries advertise themselves as offering a great degree of fine-grained control in terms of logging output, formats, and location. We take the stance that in many cases identified as "boilerplate", offering control to this degree actually is a long-term detriment. Boilerplate logging aims to move decision-making out to more coarse-grained controls to eliminate overhead that comes with subjective decision making with respect to logging outputs and formatting.

    Specifically, boilerplate logging is meant to be a to be a complement to application-specific logging, not a replacement. Our initial iteration of a boilerplate logging library written in python actually leverages use of the standard python logging library in its internal workings. The goal of boilerplate logging is remove the overhead around certain boilerplate log use cases. This allows developers to focus logging efforts on application-specific logging.

@@

    \begin{itemize}
        \item The definition and boundaries of boilerplate logging, specifically as a sub-category of logging use cases.
        \item Data and analysis of open and closed source projects implemented in python and the characterization of logging statements within.
        \item An implementation of the one of the idea presented in a Python 3+ logging library.
    \end{itemize}

@@

@@

We built a data set by analyzing various open and closed source projects of different purposes and categorizing the nature of logging statements within source code. We discovered that due to the "good citizen" nature of open source projects, most log functionality is entirely disabled or provided via hooks and left as an exercise to the caller. In fact, many open source projects perform some form of static analysis and explicitly require that logging statements considered boilerplate be left out of pull requests into the main branch. This fact correlates with our view that boilerplate logging is primarily a development activity and generally is not considered suitable for production usages. This also reinforced our understanding that boilerplate logging was in use, but the cost was felt twofold: once in the developer writing code that contains boilerplate logging statements for use in the course of development, and the second time in that developer needing to "clean up" the boilerplate logs for use in production.

@@

\begin{itemize}
    \item lines of code per file (ignoring whitespace and comments)
    \item logging lines of code per file (emitting logs or dealing with log messages, formatting, decisions, etc)
    \item boilerplate logging lines of code per file (logging lines and supporting lines that were identified to deal with boilerplate logs)
    \item expected gain ratio (overall lines of code - boilerplate lines)
\end{itemize}

@@

A second area of development cost we identify as "development time/development cycles wasted". We are unable to measure and analyze this cost quantitatively, so we took a qualitative approach. We polled a number of software developers from different backgrounds on their logging usage patterns. Specifically, the questions that we asked were:

@@

To display our ideas around the concept of boilerplate logging, we implemented a simple logging library in python, aptly named 'boilerplate'. Our initial iteration of the 'boilerplate' library includes functionality to abstract two identified "boilerplate" use cases: (a) inputs and outputs to a function or method and (b) a exception "trace" logger that caches the inputs and outputs of N calls (configurable) leading up to a thrown exception to provide more information than a typical stacktrace \footnote{Not implemented yet as part of Homework 6.}. Included with the implementation is an example python script showing the features provided by our 'boilerplate' library as well as a property-based unit test suite \footnote{Not implemented yet as part of Homework 6.}.

@@

The main entry point to our implemented logging library is as function and method decorators \footnote{See: \href{https://www.python.org/dev/peps/pep-0318/}{https://www.python.org/dev/peps/pep-0318/}} that developers can use to annotate the calls that they wish to have boilerplate logged, without any additional code required. The library uses these annotations to decide which function invocations to log as well as which channels to log. There are currently only two channel definitions: "inputs" and "outputs", and not specifying a specific channel will activate both channels by default. See Listing \ref{lst:invocation} for examples of how to trigger the boilerplate logger.

@@

Under the hood, these decorators wrap the annotated calls with code hooks into the boilerplate library, which makes decisions on what to log and with what format. Our boilerplate library leverages the use of the python standard "logging" library and exposes a lower-level "logger" object that has the same API as a built-in python "logger" object which is accessible via "boilerplate.logger" or via "logging.getLogger("boilerplate")". This provides a level of customization and familiarity that the built-in python logging library exposes, including configurations like where to log (standard out, to a file handler, etc) and what the log format should be, although the messages themselves are of a standard format.

@@

Since our library makes use of the python standard logging library and is namespaced with the name "boilerplate", developers can mix and match their application-specific logs in with the boilerplate logs or direct them to different locations. For example, a developer may find it useful to incorporate the boilerplate logs into their own specific applications on disk or in an external data store. To reduce the amount of noise, the boilerplate logger can also be disabled via standard python logging libraries API without any change in code.

@@

This registration action is exposed by our library again as a function decorator, with the syntax being "@boilerplate.register('trace')". Trace logs have both "inputs" and "outputs" channels activated for all annotated calls. Log statements are pushed to an in-memory buffer and trimmed to a configurable length in order to only keep the most relevant calls leading up to the exception.

Upon the trigger of an exception occurring in a registered function or a manual trigger by the developer via code, the logs in memory are flushed using the python standard logging library to wherever the target location is (configurable, standard out by default).

@@

We used the python static analysis tool Radon \footnote{See: \href{https://radon.readthedocs.io/en/latest/}{https://radon.readthedocs.io/en/latest/}}, which provides the four metrics we are interested in:

\begin{itemize}
    \item lines of code (logical, source, comments, etc)
    \item cyclomatic complexity (AKA McCabe complexity, number of logical decisions)
    \item maintainability index (easy to support and change, supported by Microsoft Visual Studio)
    \item various Halstead metrics\footnote{See: \href{https://en.wikipedia.org/wiki/Halstead_complexity_measures}{https://en.wikipedia.org/wiki/Halstead\_complexity\_measures}} (difficulty, effort, time to program, expected bugs, etc)
\end{itemize}

>>>>

\title{Boilerplate logger}
\subtitle{A logger for developers by developers}
\author{Evan Lee}
\affiliation{%
    \institution{University of Colorado Boulder}
    \city{Boulder}
    \state{Colorado}
}
\email{evan.n.lee@colorado.edu}
\date{May 2020}

@@

\newcommand{\code}[1]{\lstinline[basicstyle=\ttfamily\small]~#1~}

@@

\begin{abstract}
    The logging aspect of any software project always incurs a non-zero developer investment, both from a development and testing perspective as well as a operational and error tracing perspective. A significant portion of the thought and energy behind logging decisions arise from nearly-universal problems and use cases and yet still require each distinct developer to undergo the same rote process of decision making. Even with these factors and costs in play, logging is typically not considered as a first-class issue which leads to hidden cost in terms of development cycles or unnecessary iteration to assist in tracing and debugging. After having identified a widely realized use case of logging in dev-test cycles, we introduced and defined the concept of ``boilerplate logging'' and developed a Python logging library that largely abstracts the human component to logging decisions, simultaneously solving the hidden up-front development cost in logging decision making as well as eliminating any future costs in iterating on reactive postmortem logging.
\end{abstract}

@@

    % What is our contribution?
    Our contribution to this effort is two-fold: a collection and analysis of log usage in widely-used open-source libraries as well as an implementation that displays the ideas and positions presented below. Popular open-source libraries and tools are of particular interest in this space as not only do they have a wide range of contributors, but there is a communal sense of ``being a good neighbor'' when writing open-source code; particular care is given to the logging statements that are eventually committed and merged into the mainstream base and thus particular thought and effort is given to the balancing act between verbosity and utility.

@@

    In order to assist the reader in visualizing the analysis and ideas that this paper presents as well as the benefits of implementation, we will cover a code example from a widely-used Python library for interacting with SOAP interfaces: \code{python-zeep}. The code snippet can be found in Listing \ref{lst:explosion-example}\footnote{Source: \href{https://github.com/mvantellingen/python-zeep/blob/4df383021e31372c111bc26cbf2e4535deaee04e/src/zeep/transports.py\#L47}{python-zeep/transports.py}, lines 47-84. (sha: 4df383021e31372c111bc26cbf2e4535deaee04e)}. Reading through this 31-line snippet, most of the lines written by the developer deal with logging in one way or another, whether that is making logging decisions (lines 2-5, 12-22) or dealing with emitting log messages themselves (lines 6, 24). In fact, only three lines (lines 8-10) deal with any actual functionality that this method is meant to drive.

@@

    \begin{quote}
    \code{HTTP Post to https://api.colorado.edu/: \{"test": "test"\}}
    \end{quote}

@@

    \begin{quote}
        \code{[transports.post]: [args: ('https...', \{"test": "test"\}, \{\})] [kwargs: \{\}]}
    \end{quote}

@@

    The key distinction between boilerplate logging and that of other standard logging libraries is a matter of control. In general, standard third-party and language-specific logging libraries advertise themselves as offering a great degree of fine-grained control in terms of logging output, formats, and location. We take the stance that in many cases identified as ``boilerplate'', offering control to this degree actually is a long-term detriment. Boilerplate logging aims to move decision-making out to more coarse-grained controls to eliminate overhead that comes with subjective decision making with respect to logging outputs and formatting.

    Specifically, boilerplate logging is meant to be a to be a complement to application-specific logging, not a replacement. Our initial iteration of a boilerplate logging library written in Python actually leverages use of the standard Python logging library in its internal workings. The goal of boilerplate logging is remove the overhead around certain boilerplate log use cases. This allows developers to focus logging efforts on application-specific logging.

@@

    \begin{itemize}
        \item The definition and boundaries of boilerplate logging, specifically as a sub-category of logging use cases.
        \item Data and analysis of open and closed source projects implemented in Python and the characterization of logging statements within.
        \item An implementation of the idea presented in a Python 3+ logging library.
    \end{itemize}

@@

We gathered both qualitative and quantitative data with respect to logging patterns and use cases, performing analysis to uncover the niche that our identified ``boilerplate logging'' fit into over the course of software development. For our quantitative data, we performed code analysis over a set of open and closed source software projects. For our qualitative data, we conducted a survey targeted at professional software developers on personal use of logging patterns and software background.

@@

We built a data set by analyzing various open and closed source projects of different purposes and categorizing the nature of logging statements within source code. We discovered a pattern in open source software that we conjecture is due to the ``good citizen'' nature of open source projects: most log functionality is entirely disabled or provided via hooks and left as an exercise to the caller. In fact, many open source projects perform some form of linting and explicitly require that logging statements that are considered to be boilerplate be left out of pull requests into the main branch. This fact correlates with our view that boilerplate logging is primarily a development activity and generally is not considered suitable for production usages. This also reinforced our understanding that boilerplate logging was in use, but the cost was felt twofold: once in the developer writing code that contains boilerplate logging statements for use in the course of development, and the second time in that developer needing to ``clean up'' the boilerplate logs for use in production.

@@

\begin{itemize}
    \item lines of code per file (ignoring whitespace and comments)
    \item logging lines of code per file (emitting logs or dealing with log messages, formatting, decisions, etc)
    \item boilerplate logging lines of code per file (logging lines and supporting lines that were identified to deal with boilerplate logs)
    \item maximal gain ratio (<total lines of code - boilerplate lines>/<total lines>)
\end{itemize}

@@

A second area of development cost we identify as ``development time/development cycles wasted''. We were unable to measure and analyze this cost quantitatively, so we took a qualitative approach. We polled a number of software developers from different backgrounds on their logging usage patterns. Specifically, the questions that we asked were:

@@

To display our ideas around the concept of boilerplate logging, we implemented a simple logging library in Python, aptly named \code{boilerplate}. Our initial iteration of the \code{boilerplate} library includes functionality to abstract two identified ``boilerplate'' use cases: (a) inputs and outputs to a function or method and (b) a exception ``trace'' logger that caches the inputs and outputs of N calls (configurable) leading up to a thrown exception to provide more information than a typical stacktrace. Included with the implementation is an example Python script showing the features provided by our \code{boilerplate} library as well as a property-based unit test suite.

@@

The main entry point to our implemented logging library is as function and method decorators\footnote{See: \href{https://www.python.org/dev/peps/pep-0318/}{https://www.python.org/dev/peps/pep-0318/}} that developers can use to annotate the calls that they wish to have boilerplate logged, without any additional code required. The library uses these annotations to decide which function invocations to log as well as which channels to log. There are currently only two channel definitions: ``inputs'' and ``outputs'', and not specifying a specific channel will activate both channels by default. See Listing \ref{lst:invocation} for examples of how to trigger the boilerplate logger.

@@

Under the hood, these decorators wrap the annotated calls with code hooks into the boilerplate library, which makes decisions on what to log and with what format. Our boilerplate library leverages the use of the Python standard \code{logging} library and exposes a lower-level \code{logger} object that has the same API as a built-in Python \code{logger} object which is accessible via \code{boilerplate.logger} or via \code{logging.getLogger("boilerplate")}. This provides a level of customization and familiarity that the built-in Python logging library exposes, including configurations like where to log (standard out, to a file handler, etc) and what the log format should be, although the messages themselves are of a standard format.

@@

Since our library makes use of the Python standard logging library and is namespaced with the name \code{"boilerplate"}, developers can mix and match their application-specific logs in with the boilerplate logs or direct them to different locations. For example, a developer may find it useful to incorporate the boilerplate logs into their own specific applications on disk or in an external data store. To reduce the amount of noise, the boilerplate logger can also be disabled via standard Python logging libraries API without any change in code.

@@

This registration action is exposed by our library again as a function decorator, with the syntax being \code{@boilerplate.register("trace")}. Trace logs have both ``inputs'' and ``outputs'' channels activated for all annotated calls. Log statements are pushed to an in-memory buffer and trimmed to a configurable length in order to only keep the most relevant calls leading up to the exception.

Upon the trigger of an exception occurring in a registered function or a manual trigger by the developer via code, the logs in memory are flushed using the Python standard logging library to wherever the target location is (configurable, standard out by default).

@@

We made use of the tool Radon \cite{radon}, which provides the four syntactic program metrics we are interested in:

\begin{itemize}
    \item lines of code (logical, source, comments, etc)
    \item cyclomatic complexity  \cite{wiki:mccabe} (AKA McCabe complexity; number of logical decisions)
    \item maintainability index (ease of long-term support and change; supported by Microsoft Visual Studio)
    \item various Halstead metrics \cite{wiki:halstead} (difficulty, effort, time to program, expected bugs, etc)
\end{itemize}
